Continuous risk-based predictive approaches to treatment effect heterogeneity: A simulation study

### 2021-06-28 submission/manuscript.rmd

---
title: "Continuous risk-based predictive approaches to treatment effect heterogeneity: A simulation study"
abstract: |
  **Objective:** Compare risk-based methods for individualizing treatment 
  effects with simulations in the RCT setting. **Study Design and Setting:** We
  predicted absolute benefit based on an available prognostic index (PI) using:
  a model with the PI and a constant relative treatment effect; a model 
  including an interaction of treatment with the PI; 4 quarters of the PI; 
  nonlinear transformations of the PI (restricted cubic splines with 3, 4 and 5
  knots); an adaptive model selection method using Akaike’s Information 
  Criterion. Starting from a base case scenario (sample size 4,250, constant 
  odds ratio 0.8, AUC of the PI 0.75), we considered diverse assumptions by
  introducing linear and quadratic interactions of the PI with treatment and 
  varying sample size and discriminative ability of the PI. We evaluated
  performance using root mean squared error, discrimination and calibration for 
  benefit. **Results:** Models including a linear interaction of the PI with
  treatment had adequate performance that was robust under most simulation 
  scenarios. Restricted cubic splines required larger sample sizes and higher 
  AUC of the PI to achieve adequate performance. The adaptive approach performed 
  equivalently to the best-performing method in each scenario. **Conclusion:** 
  Usually, models with just a linear interaction of the PI with treatment
  adequately predict absolute benefit.
authors:
  - name: Alexandros Rekkas
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Peter R. Rijnbeek
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Ewout W. Steyerberg
    department: Department of Biomedical Data Sciences
    affiliation: Leiden University Medical Center
    location: Leiden, The Netherlands
  - name: David van Klaveren
    department: Department of Public Health
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::arxiv_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{booktabs}
  - \newcommand\given[1][]{\:#1\vert\:}
  - \date{}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

\linenumbers
# Introduction

Within the setting of patient-centered outcomes research, predictive approaches
for assessing heterogeneity of treatment effects (HTE) aim at the development of
models predicting either individualized effects or which of two (or more)
treatments is better for an individual [@Varadhan2013]. In prior work, we
divided such methods in three broader categories based on the reference class
used for defining patient similarity when making individualized predictions or
recommendations [@Rekkas2020]. Risk-modeling approaches use prediction of
baseline risk as the reference; treatment effect modeling approaches also model
treatment-covariate interactions, in addition to risk factors; optimal treatment
regime approaches focus on developing treatment assignment rules and therefore
rely heavily on modeling treatment effect modifiers.

Risk-modeling approaches to predictive HTE analyses provide a viable option in
the absence of well-established treatment effect modifiers [@Kent2019;
@PathEnE]. In simulations, modeling of effect modifiers in the form of
treatment-covariate interactions often led to miscalibrated predictions of
benefit, while risk-based methods proved quite robust [@vanKlaveren2019]. Most
often, risk-modeling approaches are carried out in two steps: first a risk
prediction model is developed externally or internally on the entire RCT
population, "blinded" to treatment; then the RCT population is stratified using
this prediction model to evaluate risk-based treatment effect variation
[@Kent2010]. However, even though estimates at the risk subgroup level are
accurate, this does not apply on the individual level, especially for patients
with predicted risk at the boundaries of the risk intervals. Therefore, the
risk-stratified approach should be used for exploring and presenting an overview
of HTE, while inferences on the individual level should be made with caution.

We aimed to provide an overview of methods that can be used to move from a
risk-stratified approach to a continuous one using common smoothing techniques.
These methods extend the risk-based framework of predictive HTE analyses to
allow predictions on the individual level, within the RCT setting. We carried
out a simulation study to compare the performance of these methods under
different settings of increasing non-linearity of treatment effects. Finally, we
carried out an application on real data as a demonstration of the considered
techniques.

@@

The linear interaction model proved to be flexible enough to adequately predict
benefit under a wide range of scenarios. It had lower error rates and usually
the highest c-for-benefit with the lower variability across simulation
replications. Its benefit predictions were also well-calibrated across the
majority of the simulation settings. Models with RCS smoothing with 3 knots
performed very similarly, only outperforming the linear interaction model in the
presence of strong quadratic deviations from the base case scenario of constant
relative treatment effects. These simpler methods proved to have quite robust
performance across all scenarios, which was not the case for more flexible
approaches.

Methods with RCS smoothing with 4 or 5 knots proved to be data-hungry and this
often led to overfitting with smaller sample sizes. This resulted in higher
RMSE, increased variability of discrimination for benefit and worse-calibrated
benefit predictions across simulation replications. Even with larger sample
sizes and strong quadratic deviations from the base case scenario of constant
relative treatment effects, these more flexible RCS smoothing methods did not
outperform simpler RCS smoothing with 3 knots. More flexible approaches usually
do not come with any added value, unless we expect extreme heterogeneity in
treatment effects.

The constant treatment effect model, despite having adequate performance even in
the presence of weak treatment effect heterogeneity on the relative scale, it
quickly broke down with stronger deviations from the base case scenario of
constant effects. In these cases, the stratified approach generally had lower
error rates compared to the constant treatment effect model. In concept, the
stratified approach lies between the constant effects model and smoother
approaches, only assuming constant treatment effects within strata of predicted
risk and, therefore, is more sensitive to treatment effect heterogeneity.
However, these stepwise benefit estimates can only be used as a demonstration of
treatment effect heterogeneity, identifying patient subgroups where treatment
benefits are expected to be higher or lower, and should not be used to make
individualized absolute benefit predictions.

Increasing the true prediction AUC by increasing the predictor coefficients of
the true risk model reduced RMSE for all methods, similar to increasing the
sample size. Higher AUC translates in higher variability of predicted risks,
which, in turn, allows the considered methods to better follow the evolution of
treatment benefit. Higher prediction AUC also led to higher absolute values of
c-for-benefit. Even though, model performance appears to be very important for
the performance of these methods, prediction model development was outside the
scope of this work and has already been studied quite extensively
[@vanKlaveren2019; @Burke2014; @Abadie2018].

Risk-based approaches to predictive HTE assume that benefit is a function of
baseline risk. A limitation of our study is that this was assumed to be the case
in the majority of the simulation scenarios. Even though our main conclusions
did not change when simulating true treatment-covariate interactions (results in
supplement), RMSE had higher values. The linear interaction model under these
scenarios again had the best overall performance under different sizes of the
interaction effects. This can be explained by the fact that linear interactions
with treatment were simulated, thus favoring the linear interaction model. In
studies evaluating prediction model development, the underlying outcome
generation process sometimes proved to affect performance of certain methods
[@Austin2021]. This might be the case here as well and should be explored in
more extensive future simulation studies.

@@

In conclusion, when comparing different risk-based approaches to individualizing
benefit predictions simpler models including a linear interaction with the
prognostic index or RCS smoothing with 3 knots had adequate performance under a
wide range scenarios. An adaptive approach selecting among the considered
methods with optimal AIC had satisfactory performance.

>>>>

---
title: "Continuous risk-based predictive approaches to treatment effect heterogeneity: A simulation study"
abstract: |
 **Objective:** Baseline risk is an important determinant of the absolute effect of a treatment for an individual patient. We aimed to compare different risk-based methods for predicting individualized treatment effects with simulations of the RCT setting. **Study Design** and Setting: We sampled patients from a superpopulation with diverse assumptions for a baseline prognostic index of risk (PI), and for the shape of the interaction between the PI and treatment (no, linear or quadratic interaction). In each sample, we fitted different models for predicting absolute benefit: a model with the PI and a constant relative treatment effect and models including an interaction of treatment assignment with the PI, with 4 quarters of the PI, and with nonlinear transformations of the PI (restricted cubic splines with 3, 4 and 5 knots). We also used an adaptive model selection approach based on Akaike’s Information Criterion. We evaluated predictive performance in the superpopulation using root mean squared error and measures of discrimination and calibration for benefit. Starting from a base case scenario (sample size 4,250, treatment odds ratio 0.8, AUC of the PI 0.75), we varied the sample size,  the strength of the treatment effect, and the discriminative ability of the PI. **Results:** Models including a linear interaction of the PI with treatment had (close to?) optimal performance under most simulation scenarios. More flexible models with restricted cubic splines required larger sample sizes and higher AUC of the PI to outperform the linear model. The adaptive approach performed similarly to the best-performing method in each scenario. **Conclusion:** Under most circumstances, a model with a linear interaction of the PI with treatment is better able to predict absolute treatment benefit.
authors:
  - name: Alexandros Rekkas
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Peter R. Rijnbeek
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Ewout W. Steyerberg
    department: Department of Biomedical Data Sciences
    affiliation: Leiden University Medical Center
    location: Leiden, The Netherlands
  - name: David van Klaveren
    department: Department of Public Health
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::arxiv_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{booktabs}
  - \newcommand\given[1][]{\:#1\vert\:}
  - \date{}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

\linenumbers
# Introduction
Predictive approaches for assessing heterogeneity of treatment effects (HTE) aim
at the development of models predicting either individualized effects or which
of two (or more) treatments is better for an individual [@Varadhan2013]. In
prior work, we divided such methods in three broader categories based on the
reference class used for defining patient similarity when making individualized
predictions or recommendations [@Rekkas2020]. Risk-modeling approaches use
prediction of baseline risk as the reference; treatment effect modeling
approaches also model treatment-covariate interactions, in addition to risk
factors; optimal treatment regime approaches focus on developing treatment
assignment rules and therefore rely heavily on modeling treatment effect
modifiers.

Risk-modeling approaches to predictive HTE analyses provide a viable option in
the absence of well-established treatment effect modifiers [@Kent2019;
@PathEnE]. In simulations, modeling of effect modifiers, i.e.
treatment-covariate interactions, often led to miscalibrated predictions of
benefit, while risk-based methods proved quite robust [@vanKlaveren2019]. Most
often, risk-modeling approaches are carried out in two steps: first a risk
prediction model is developed externally or internally on the entire RCT
population, “blinded” to treatment; then the RCT population is stratified using
this prediction model to evaluate risk-based treatment effect variation
[@Kent2010]. However, even though estimates at the risk subgroup level may be
accurate, these estimated do not apply to individual patients, especially for
patients with predicted risk at the boundaries of the risk intervals. Hence, the
risk-stratified approach is useful for exploring and presenting HTE, but is not
useful for supporting treatment decisions for individual patients.

To individualize treatment effects, the recent PATH statement suggested various
risk-based models including a prognostic index of baseline risk (PI) and
treatment assignment [cite PATH !!]. We aimed to summarize and compare different
risk-based models for predicting individualized treatment effects. We simulated
RCT settings to compare the performance of these models under different
assumptions of the relationship between baseline risk and treatment. We
illustrated the different models by a case study of predicting individualized
effects of tissue plasminogen activator (tPA) versus streptokinase treatment in
patients with an acute myocardial infarction (MI).

@@

The linear interaction model proved to be superior to alternative approaches
when predicting risk-based treatment benefit under a wide range of scenarios. It
generally had lower mean squared error, lower e-for-benefit and higher
c-for-benefit with lower variability across simulation replications. Models
including restricted cubic splines with 3 knots only outperformed the linear
interaction model in the presence of strong quadratic deviations from a constant
relative treatment effect.

Including restricted cubic splines with 4 or 5 knots proved to be too flexible,
as indicated by higher RMSE, increased variability of discrimination for benefit
and worse calibration of benefit predictions. Even with larger sample sizes and
strong quadratic deviations from the base case scenario of constant relative
treatment effects, these more flexible restricted cubic splines did not
outperform the simpler RCS with 3 knots. These more flexible approaches may only
be helpful if we expect more extreme patterns of heterogeneous treatment
effects.

The constant treatment effect model, despite having adequate performance in the
presence of weak treatment effect heterogeneity on the relative scale, quickly
broke down with stronger deviations from constant relative treatment effects. In
these cases, the stratified approach generally had lower error rates compared to
the constant treatment effect model. In concept, the stratified approach lies
between the constant effects model and smoother approaches, only assuming
constant treatment effects within strata of predicted risk and, therefore, is
more sensitive to treatment effect heterogeneity. Stepwise treatment benefit
estimates are very useful for demonstrating treatment effect heterogeneity –
because estimating treatment effect requires groups of patients rather than
individual patients –but are not helpful for making individualized absolute
benefit predictions.

Increasing the discriminative ability of the risk model – by increasing the
predictor coefficients of the true risk model – reduced RMSE for all methods,
similar to increasing the sample size. This increase in discriminative ability
translates in higher variability of predicted risks, which, in turn, allows the
considered methods to better follow the evolution of treatment benefit. As
expected, the increase in discriminative ability of the risk model also led to
higher values of c-for-benefit. Even though risk model performance is very
important for the ability of risk-based methods to predict treatment benefit,
prediction model development was outside the scope of this work and has already
been studied extensively [@vanKlaveren2019; @Burke2014; @Abadie2018].

Risk-based approaches to predictive HTE estimate treatment benefit as a function
of baseline risk. A limitation of our study is that we assumed treatment benefit
to be a function of baseline risk in the majority of the simulation scenarios.
Nevertheless, our main conclusions did not change when we generated individual
treatment-covariate interactions (supplementary table/figure x). Future
simulation studies could explore the effect of more extensive deviations from
risk-based treatment effects.

@@

In conclusion, when comparing different risk-based approaches to predict
individualizing benefit a models including a linear treatment interaction with
the prognostic performed best in a wide range of scenarios. More flexible models
with restricted cubic splines required larger sample sizes and higher AUC of the
PI to outperform the linear model. An adaptive approach, selecting the model
with optimal AIC, had satisfactory performance.
